---
title: "The modified fpca.face function in refund"
author: "Ruonan Li"
header-includes: 
  - \usepackage{bm}
  - \usepackage{amsfonts}
  - \usepackage{amsthm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{mathrsfs}
  - \usepackage{array}
  - \usepackage{times}
  - \usepackage{dsfont}
  - \usepackage{caption}
  - \usepackage{multirow}
output: pdf_document
---


# 1. Algorithm.
For subject $i= 1,...,I$ denote by $Y_i(t)$ the observed data at $t \in \mathcal{T}$. Assume that data are observed on a regular grid of the domain, that is $\{t_{1},...,t_{J}\}$, where $J$ is the total number of grid points. Define $\mathbf{Y}$ the demeaned observed functional data, which is a $J\times I$ dimensional matrix with the $i$th column equal to $\mathbf{Y}_i$.


**Part 1: FACE algorithm in Xiao et al. (2016). **

\begin{itemize}
\item Step 1: conduct the  eigendecomposition $(\mathbf{B}^T\mathbf{B})^{-1/2}\mathbf{P}(\mathbf{B}^T\mathbf{B})^{-1/2}=\mathbf{U}\mathbf{D}\mathbf{U}^T $, where $\mathbf{U}$ is an orthonormal matrix and $\mathbf{D}$ is a diagonal matrix.
\item Step 2: calculate $\mathbf{A}_{\mathbf{S}}=\mathbf{B}(\mathbf{B}^T\mathbf{B})^{-1/2}\mathbf{U}$.
\item Step 3: calculate $\tilde{\mathbf{Y}} = \mathbf{A}^T_{\mathbf{S}}\mathbf{Y}$.
\item Step 4: conduct the eigendecomposition $ I^{-1}(\mathbf{I}_c+\lambda\mathbf{D})^{-1} \tilde{\mathbf{Y}}\tilde{\mathbf{Y}}^T (\mathbf{I}_c+\lambda\mathbf{D})^{-1} = \mathbf{A}\mathbf{\Sigma}\mathbf{A}^T$, where $\mathbf{A}$ is an orthonormal matrix and $\mathbf{\Sigma}$ is a diagonal matrix.
\item Step 5: calculate eigenvetors of $\tilde{\mathbf{K}}$ as $\mathbf{A}_{\mathbf{S}}\mathbf{A}$.
\end{itemize}




**Part 2: The modified FACE algorithm. **

\begin{itemize}
\item Step 1: conduct the  eigendecomposition $(\mathbf{B}^T\mathbf{B})^{-1/2}\mathbf{P}(\mathbf{B}^T\mathbf{B})^{-1/2}=\mathbf{U}\mathbf{D}\mathbf{U}^T $, where $\mathbf{U}$ is an orthonormal matrix and $\mathbf{D}$ is a diagonal matrix.
\item Step 2: calculate  $\mathbf{A}_{0}=(\mathbf{B}^T\mathbf{B})^{-1/2}\mathbf{U}$ and $\mathbf{A}_{\mathbf{S}}=\mathbf{B}(\mathbf{B}^T\mathbf{B})^{-1/2}\mathbf{U}$.
\item Step 3: calculate $\tilde{\mathbf{Y}} = \mathbf{A}^T_{\mathbf{S}}\mathbf{Y}$.
\item Step 4: construct transformed  basis functions as $\tilde{\mathbf{B}}(s)$ so that they are orthonormal, i.e., $\int \tilde{\mathbf{B}}(s)\tilde{\mathbf{B}}^T(s) ds = \mathbf{I}_c$. We have $\tilde{\mathbf{B}} = \mathbf{B}\mathbf{G}^{-1/2}$ and $\tilde{\mathbf{A}}_{0}=\mathbf{G}^{1/2}\mathbf{A}_{0}$, where $\mathbf{G}=\int \mathbf{B}(s)\mathbf{B}^T(s)ds$.
\item Step 5: conduct the eigendecomposition $\tilde{\mathbf{A}}_{0}\{ n^{-1}(\mathbf{I}_c+\lambda\mathbf{D})^{-1} \tilde{\mathbf{Y}}\tilde{\mathbf{Y}}^T (\mathbf{I}_c+\lambda\mathbf{D})^{-1}\}\tilde{\mathbf{A}}^T_{0} = \mathbf{A}\mathbf{\Sigma}\mathbf{A}^T$, where $\mathbf{A}$ is an orthonormal matrix and $\mathbf{\Sigma}$ is a diagonal matrix.
\item Step 6: The  estimated $k$th  eigenfunction is $\boldsymbol{\phi}_k(s) = \mathbf{A}^T_k\tilde{\mathbf{B}}(s)$, where $\mathbf{A}_k$ is the $k$th column vector of $\mathbf{A}$.
\end{itemize}


We use mixed model equations (MME) to estimate principal scores. Assume that we obtained ${Y}_{i}(t)$, eigenvalues $\lambda_{k}$, and eigenfunctions $\phi_{k}(t), k = 1, \dots, N$ from previous steps. $N$ is selected using the proportion of explained variance. The mixed model is
\begin{equation}
    \begin{cases}
        {Y}_{i}(t) = \sum_{k = 1}^{N}\xi_{ik}\phi_{k}(t) + \epsilon_{i}(t)\;,\\
        \xi_{ik} \sim \mathcal{N}\{0, \lambda_{k}\}\;; \quad \epsilon_{i}(t) \sim \mathcal{N}(0, \sigma^2)\;.
    \end{cases}  \label{eq:mixedmodelsubj}     
\end{equation}

Data may be incompletely collected, that is $\{t_{1},...,t_{T_i}\}$. Let ${\mathbf{Y}}_{i} = ({Y}_{i1},\cdots, {Y}_{iT_i})^T$ and $\boldsymbol{\xi}_{i} = (\xi_{i1},\cdots, \xi_{iN})^T$. Note that the score $\boldsymbol{\xi}_{i}$ has a diagonal covariance matrix $\boldsymbol{\Sigma} = \text{diag}(\lambda_{1},\cdots, \lambda_{N})$. Let $\boldsymbol{\phi}_{k,i} = (\phi_{k}(t_{1}),\cdots, \phi_{k}(t_{T_i}))^T$ and $\boldsymbol{\Phi}_i = [\boldsymbol{\phi}_{1,i},\cdots, \boldsymbol{\phi}_{N,i}] \in \mathbb{R}^{T_i\times N}$. Let $\boldsymbol{\epsilon}_{i}=(\epsilon_{i1},\cdots, \epsilon_{iT_i})^T$. Then
$$
\mathbf{Y}_{i} = \boldsymbol{\Phi}_i\boldsymbol{\xi}_{i} +\boldsymbol{\epsilon}_{i}\;.
$$
By MME, we have 
\begin{equation} \label{MME}
\boldsymbol{\xi}_{i} = \{\boldsymbol{\Phi}_i^T\boldsymbol{\Phi}_i + \sigma^2\Sigma^{-1}\}^{-1}\boldsymbol{\Phi}_i^T{\mathbf{Y}}_{i},
\end{equation}
where the dimension of the matrix to take inverse is $N \times N$. Then,  $\hat{\mathbf{Y}}_{i} = \boldsymbol{\Phi}_i\boldsymbol{\xi}_{i}$.


# 2. Simulation: compare the computation time and estimate accuracy of `fpca.face` and `fpca.face2`.

## 2.1 Equally-spaced data.
We assume that the regular functional data are observed on an equally-spaced grid $\{t_1, ..., t_J\}$ of domain $\mathcal{T} = [0, 1]$ such that $t_j= j/J$ for $j = 1, \dots, J$. We also consider an incomplete scenario, where the number of observed points per function is $0.5\times J$. We considered $I=100, 200, 500$ subjects, and $J = 100, 1000, 10000$ points per curve. The magnitude of noise $\sigma = 0.5$. 


We use a similar simulation setting with Xiao et al. (2016). Consider the following model
\begin{align*}
    {Y}_{i}(t) = \sum_{k = 1}^{3}\xi_{ik}\phi_{k}(t) + \epsilon_{i}(t)\;,
\end{align*}
where $\xi_{ik} \sim \mathcal{N}\{0, \lambda_{k}\}$ and $\epsilon_{i}(t) \sim \mathcal{N}(0, \sigma^2)$. The true eigenvalues are $\lambda_{k}=0.5^{k-1}, k=1,2,3$. The true eigenfunctions are selected as $\phi_{k}(t) = \{1, \sqrt{3}(2t-1), \sqrt{5}(6t^2 - 6t + 1)\}$. Those eigenfunctions are not mutually orthogonal.


I simulated $100$ data sets for each scenario and compared `fpca.face` and `fpca.face2` by computation time, mean integraed square errors (MISE) for eigenfunctions and mean square errors (MSE) for estimated scores. Results are shown in Table \ref{tab:equal}. 


```{r equal, echo=FALSE, warning=FALSE, message=FALSE}
# Load necessary libraries
library(knitr)
library(kableExtra)
library(dplyr)

# Read the CSV file
results_eq <- read.csv("sim/simulation_results_equal_TRUE.csv")

results_eq %>%
  select(design, n, everything(), -X) %>%
  kbl(booktabs = T, digits = 4, format = "latex", escape = F,
      col.names = c("deisgn", "J", "n", "time (s)", "$MISE(\\phi)$", "MSE",
                    "time (s)", "$MISE(\\phi)$", "MSE"),
      caption = "Equally-spaced data: the average MISE for eigenfunctions and the average MSE for scores across $100$ simulations.") %>%
  add_header_above(c(
    " " = 3, 
    "fpca.face" = 3, 
    "fpca.face2" = 3
  )) %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  collapse_rows(columns = 1:3, valign = "middle", latex_hline = "custom", 
                custom_latex_hline = 1:2)
```


## 2.2 Unequally-spaced data.
For unequally-spaced data, data are also observed on $J$ points. The $J$ points are randomly sampled from an equally-spaced grid $\{t_1, ..., t_{5J}\}$ of domain $\mathcal{T} = [0, 1]$. Results are shown in Table \ref{tab:unequal}. 

```{r unequal, echo=FALSE}
# Load necessary libraries
library(knitr)
library(kableExtra)
library(dplyr)

# Read the CSV file
results_neq <- read.csv("sim/simulation_results_equal_FALSE.csv")

results_neq %>%
  select(design, n, everything(), -X) %>%
  kbl(booktabs = T, digits = 4, format = "latex", escape = F,
      col.names = c("deisgn", "J", "n", "time (s)", "$MISE(\\phi)$", "MSE",
                    "time (s)", "$MISE(\\phi)$", "MSE"),
      caption = "Unequally-spaced data: the average MISE for eigenfunctions and the average MSE for scores across $100$ simulations.") %>%
  add_header_above(c(
    " " = 3, 
    "fpca.face" = 3, 
    "fpca.face2" = 3
  )) %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  collapse_rows(columns = 1:3, valign = "middle", latex_hline = "custom", 
                custom_latex_hline = 1:2)
```





